---
title: "STATS19 reported road casualties: LA region of `r params$authority`"
date: today
params:
  authority: "Bristol"
  output_dir: "outputs/Bristol"
format:
  html:
    toc: true
    toc-location: left
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(stats19reports)
library(dplyr)
library(sf)

la_dir <- params$output_dir

# all objects created by build_la_report_data()
dat <- readRDS(file.path(la_dir, "data", "la_report_data.rds"))
list2env(dat, environment())

n_las <- NROW(LAs)

# derived stats for inline text
osm   <- report_osm_stats(cas_osm_period, cas_osm_year, cas_osm_type)
spd   <- report_speed_stats(csl)
imd   <- report_imd_stats(imd_casualties)
demog <- report_demog_stats(age_sex)
msoa  <- report_msoa_stats(cas_df_all_LA)
pave_la <- report_pavement_stats(sing_veh_pave)
pave_gb <- report_pavement_stats(sing_veh_pave_gb)

fig <- function(...) knitr::include_graphics(file.path(la_dir, ...))
```

# Introduction

This is a report for the Local Authority (LA) region of `r la_name`, which
can be reproduced for any LA with the `stats19reports` package (see
<https://github.com/BlaiseKelly/stats19reports>).

In this report all serious and slight casualty data is based on the adjusted
data.

`r la_summary_paragraph(LA_LY, LA_YBLY, n_las, YBLY, la_name)`

```{r}
#| layout-ncol: 3
fig("plots/national/la_casualties_All_fatal_cas_LY.png")
fig("plots/national/la_casualties_All_ksi_cas_LY.png")
fig("plots/national/la_casualties_All_total_cas_LY.png")
```

`r five_year_sentence(LA_LY, LA_5Y)`

```{r}
fig("plots/national/rank_All.png")
```

`r la_summary_paragraph(LA_LY_CYC, LA_YBLY_CYC, n_las, YBLY, la_name, casualty_label = "cyclist")`

```{r}
#| layout-ncol: 3
fig("plots/national/la_casualties_Cyclist_fatal_cas_LY.png")
fig("plots/national/la_casualties_Cyclist_ksi_cas_LY.png")
fig("plots/national/la_casualties_Cyclist_total_cas_LY.png")
```

`r five_year_sentence(LA_LY_CYC, LA_5Y_CYC, casualty_label = "cyclists")`

```{r}
fig("plots/national/rank_Cyclist.png")
```

`r la_summary_paragraph(LA_LY_PED, LA_YBLY_PED, n_las, YBLY, la_name, casualty_label = "pedestrian")`

```{r}
#| layout-ncol: 3
fig("plots/national/la_casualties_Pedestrian_fatal_cas_LY.png")
fig("plots/national/la_casualties_Pedestrian_ksi_cas_LY.png")
fig("plots/national/la_casualties_Pedestrian_total_cas_LY.png")
```

`r five_year_sentence(LA_LY_PED, LA_5Y_PED, casualty_label = "pedestrians")`

```{r}
fig("plots/national/rank_Pedestrian.png")
```

## Where in `r la_name`?

The interactive maps saved in `` `r file.path(la_dir, "maps")` `` show the
location of all collisions between `r base_year` and `r upper_year`, sized
by severity of the casualties and coloured by variables such as the speed
limit of the road, with radio buttons to disaggregate by casualty type.
Clicking on each casualty provides further information. Grouping by street
using OpenStreetMap data, the OSM maps show the collisions for each street
coloured by KSI and disaggregated by year; each street can be clicked for a
pop-up with more details.

## Road Network Analysis

Casualty data was matched to `r osm$n_links` road links from OpenStreetMap
for the period `r base_year` to `r upper_year`, covering
`r osm$total_collisions` collisions. The road with the highest number of
collisions was `r osm$top_road_name` (`r osm$top_road_collisions`), followed
by `r osm$second_road_name` (`r osm$second_road_collisions`). Across
`r osm$n_years` years the peak year for collisions on mapped roads was
`r osm$peak_year` with `r osm$peak_year_n` collisions. By casualty type the
most common on the OSM network was `r osm$top_type` (`r osm$top_type_n`
collisions). The table below shows roads by collision count for the full
period. Click a header to sort the column.

```{r}
tabulate_osm_roads(cas_osm_period,
                   save_path = file.path(la_dir, "tables", "cas_osm.html"))
```

The plots below show collisions on the 2 roads with the most collisions in
`r la_name`: `r osm$top_road_name` and `r osm$second_road_name`. The plot on
the left represents the casualties and the plot on the right vehicles
involved. To see a plot larger, right click and either save or "open in new
tab".

```{r}
#| layout-ncol: 2
n1_rd <- gsub(" ", "_", osm$top_road_name)
fig("plots/streets", paste0(n1_rd, "_cas_map.png"))
fig("plots/streets", paste0(n1_rd, "_All_veh_map.png"))
```

```{r}
#| layout-ncol: 2
n2_rd <- gsub(" ", "_", osm$second_road_name)
fig("plots/streets", paste0(n2_rd, "_cas_map.png"))
fig("plots/streets", paste0(n2_rd, "_All_veh_map.png"))
```

## Speed Limits: Casualty rates by road speed limit

In `r la_name`, according to OSM data the majority of roads are
`r spd$majority_limit`mph (`r spd$majority_pc`%). Between `r base_year` and
`r upper_year` the majority of collisions (`r spd$max_coll`) happened on
`r spd$max_coll_limit`mph roads. `r spd$fatal_pc_at_max`% of fatalities
happened on `r spd$max_fatal_limit`mph roads, which make up
`r spd$pc_length_30`% of roads in the LA. `r spd$max_serious` casualties
were seriously injured on `r spd$max_serious_limit`mph roads. Collision
rates per km are highest for `r spd$max_coll_km_limit`mph roads, fatality
rates highest on `r spd$max_fatal_km_limit`mph roads, serious injury rates
highest on `r spd$max_serious_km_limit`mph and slight injury rates
`r spd$max_slight_km_limit`mph. The collision rate on 20mph roads was
`r spd$coll_km_20_vs_max`% of the highest rate
(`r spd$max_coll_km_limit`mph roads), and fatality rates
`r spd$fatal_km_20_vs_30`% of 30mph roads and `r spd$fatal_km_20_vs_max`%
of the highest fatality rate (`r spd$max_fatal_km_limit`mph) roads. KSI
rates were highest on `r spd$max_ksi_km_limit`mph roads,
`r spd$ksi_km_30_vs_20` times higher on 30mph than 20mph roads.

The chart below shows percentage of fatalities by road speed limit.

```{r}
fig("plots/conditions/speed_limit_fatal.png")
```

Applying the 30mph fatality rate (`r spd$fatal_col_30` fatalities per
collision) to the number of 20mph collisions would give
`r round(spd$fatal_col_30 * spd$collisions_20)` fatalities. Applying the
30mph per-km fatality rate to the 20mph road network would give
`r round(spd$fatal_km_30 * spd$tot_length_20)`
fatalities.

```{r}
tabulate_speed_limits(csl, la_name, base_year, upper_year,
                      save_path = file.path(la_dir, "tables", "cas_col_rate"))
```

## Grouping by area and IMD

### LSOA

The top left plot below shows the distribution of the population by LSOA
zone,
[the most granular population data available for the whole of Great Britain](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/lowersuperoutputareamidyearpopulationestimates).
In the top right are the number of total collisions by LSOA zone, calculated
by intersecting the crash data with the LSOA shape files, and the bottom two
plots show the home LSOA of casualties and vehicles involved in collisions
in the city.

```{r}
fig("plots/lsoa/lsoa_all.png")
```

### Relationship with Index of Multiple Deprivation Decile (IMD)

The Index of Multiple Deprivation (IMD) is the official measure of relative
deprivation in England and is part of a suite of outputs that form the
Indices of Deprivation (IoD), given as a decile in the STATS19 casualty and
vehicle data for the casualty and driver's home LSOA respectively. It
follows an established methodological framework in broadly defining
deprivation to encompass a wide range of an individual's living conditions.
It is important in road traffic collision analysis because it says something
about where harm is distributed in society. In `r la_name`,
`r ifelse(pop_least_imd > 50, pop_least_imd, 100 - pop_least_imd)`% of the
population fall into the
`r ifelse(pop_least_imd > 50, "least", "most")` deprived IMD deciles.
`r ifelse(cas_least_imd > 50, cas_least_imd, 100 - cas_least_imd)`% of
casualties injured between `r base_year` and `r upper_year` fall into the
`r ifelse(cas_least_imd > 50, "least", "most")` deprived IMD deciles.
`r cas_imd_data`% of casualty data had home LSOA.

### IMD and Casualties

The relationship between deprivation and road casualty risk is marked in
`r la_name`. The most deprived IMD decile (decile 1) accounts for
`r imd$d1_pc`% of casualties (`r imd$d1_n`), while the least deprived
(decile 10) accounts for `r imd$d10_pc`% (`r imd$d10_n`). Residents of the
most deprived areas are `r imd$ratio` times more likely to be a casualty
than those in the least deprived decile. The decile with the highest number
of casualties overall is decile `r imd$peak_decile` with `r imd$peak_n`
casualties (`r imd$peak_pc`%). As noted above, `r cas_imd_data`% of casualty
records had a home LSOA that could be matched to an IMD decile.

The charts below show a breakdown of the ages, sex, number of KSI and IMD
decile of casualties in `r la_name`.

```{r}
#| layout-ncol: 1
fig("plots/imd/ksi_imd_age.png")
fig("plots/imd/ksi_imd_sex.png")
```

Analysis of KSI casualties by IMD decile in `r la_name` reveals that decile
`r imd$top_imd$IMDDecil`
(`r ifelse(imd$top_imd$IMDDecil <= 3, "most deprived", ifelse(imd$top_imd$IMDDecil >= 8, "least deprived", "mid-range"))`)
accounts for the highest KSI total across casualty types combined, with
approximately `r round(imd$top_imd$ksi)` KSI casualties.
`r imd$top_type$casualty_type`s represent the highest KSI casualty type
overall (~`r round(imd$top_type$ksi)`). Pedestrian KSI peaks in decile
`r imd$top_imd_type$IMDDecil[imd$top_imd_type$casualty_type == "Pedestrian"]`,
while cyclist KSI is highest in decile
`r imd$top_imd_type$IMDDecil[imd$top_imd_type$casualty_type == "Cyclist"]`.
By age, the most affected groups are
`r imd$top_age_type$dft_age_band[imd$top_age_type$casualty_type == "Pedestrian"]`
for pedestrians,
`r imd$top_age_type$dft_age_band[imd$top_age_type$casualty_type == "Cyclist"]`
for cyclists, and
`r imd$top_age_type$dft_age_band[imd$top_age_type$casualty_type == "Motorcyclist"]`
for motorcyclists. Males account for approximately
`r round(imd$sex_split$ksi[imd$sex_split$sex_of_casualty == "Male"])` KSI
casualties compared to
`r round(imd$sex_split$ksi[imd$sex_split$sex_of_casualty == "Female"])` for
females.

### MSOA

Between LSOA and Local Authority are Middle Layer Super Output Areas
(MSOA). Because of their relatively consistent size across the country, they
are a useful unit for analysing data about small local areas, and they have
also been allocated descriptive names by
[the House of Commons Library](https://houseofcommonslibrary.github.io/msoanames/),
making it easier to describe areas. These areas have been used to analyse
casualties by MSOA region and compare at a country level, similar to what
has been done with LAs at the beginning of this document, but hopefully
providing more targeted information for a local area.

Analysing the population weighted IMD from LSOA for MSOA for different
casualty types, the strongest trends are for
`r paste(unique(arrange(slope_dat_la, desc(r2))$casualty_type[1:8]), collapse = ", ")`
casualties as shown in the plot below.

```{r}
fig("plots/msoa/la_8.png")
```

The table below shows data for the `r msoa$n_msoa` MSOA areas in `r la_name`
where casualties live, ranked against other MSOAs nationally. Casualties
involved in collisions outside of `r la_name` are also included. Between
`r base_year` and `r upper_year`, the MSOA with the highest number of
casualties was `r msoa$top_msoa` with `r round(msoa$top_msoa_n)` casualties.
The most deprived MSOA (by weighted IMD score) was
`r msoa$most_deprived_msoa` (IMD `r msoa$most_deprived_imd`), while the
least deprived was `r msoa$least_deprived_msoa` (IMD
`r msoa$least_deprived_imd`). The highest ranked MSOA regions per capita
where casualties live in `r la_name` are `r msoa$top_fatal_pcap_msoa`,
ranked `r msoa$top_fatal_pcap_rank` nationally for fatalities per capita,
`r msoa$top_ksi_pcap_msoa` ranked `r msoa$top_ksi_pcap_rank` for KSIs, and
`r msoa$top_serious_pcap_msoa` ranked `r msoa$top_serious_pcap_rank` for
serious casualties.

```{r}
tabulate_msoa_ranks(cas_df_all_LA,
                    save_path = file.path(la_dir, "tables",
                                          "cas_msoa_rank.html"))
```

## Casualty Demographics

Between `r base_year` and `r upper_year`, `r demog$male_pc`% of all
casualties in `r la_name` were male and `r demog$female_pc`% were female.
Males also accounted for `r demog$male_ksi_pc`% of KSI casualties. The most
common age band for casualties was `r demog$peak_age` (`r demog$peak_age_pc`%
of all casualties), while the highest KSI count was among the
`r demog$peak_ksi_age` age band (`r demog$peak_ksi_age_pc`% of all KSI). The
table below shows the full breakdown of casualties by age band and sex, with
severity and the percentage share of KSI and all casualties.

```{r}
fig("plots/demog/demographics.png")
```

```{r}
tabulate_age_sex(age_sex)
```

## Single Vehicle Pavement Collisions

Between `r base_year` and `r upper_year`, nationally there were
`r pave_gb$total_collisions` single-vehicle collisions where a pedestrian
was struck on a footway or verge. `r pave_gb$motor_coll_pc`% involved people
driving cars, taxis, goods vehicles and buses and `r pave_gb$micro_coll_pc`%
involved people riding pedal cycles and e-scooters. Of these collisions
there were `r pave_gb$total_fatal` fatalities; `r pave_gb$motor_fatal_pc`%
of the fatal collisions were with people driving cars, taxis, goods vehicles
and buses and `r pave_gb$micro_fatal_pc`% with people riding pedal cycles
and e-scooters. There were a further `r pave_gb$total_serious` serious
injuries, of which `r pave_gb$motor_serious_pc`% were with people driving
cars, taxis, goods vehicles and buses and `r pave_gb$micro_serious_pc`% with
people riding pedal cycles and e-scooters.

In `r la_name` there were `r pave_la$total_collisions` reported incidents,
resulting in `r pave_la$total_cas` casualties. The vehicle type most
commonly involved was `r pave_la$top_type` with `r pave_la$top_coll`
collisions and `r pave_la$top_n` casualties. `r pave_la$motor_coll_pc`% of
the vehicles in collisions were driven cars, taxis, goods vehicles and buses
and `r pave_la$micro_coll_pc`% were ridden pedal cycles and e-scooters. Of
these collisions there were `r pave_la$total_fatal` fatalities and
`r pave_la$total_serious` serious injuries, of which
`r pave_la$motor_serious_pc`% were with people driving cars, taxis, goods
vehicles and buses and `r pave_la$micro_serious_pc`% with people riding
pedal cycles and e-scooters.

The plot below visually represents a pavement, with squares for all the KSI
casualties in `r la_name` coloured by the vehicle that hit them.

```{r}
fig("plots/streets/KSI_pavements.png")
```

## Appendix

### Casualty type categorisation

```{r}
tabulate_casualty_groupings(cas_type)
```
